Making Sequential Programs Deterministic

### 2019-03-29 design.tex
todo: collection types

Figure~\ref{typecheck-rules-standard} shows a sample of the standard typing rules applied to 
any typed language. In the generic case, the type $\tau$ represents a valid basetype in a programming language.
\ourTypeSystem extends a basetype with a type qualifier. Throughout this paper, $\tau$ represents a fully qualified type.
In other words, type $\tau = \kappa \ \beta$ consists of both a qualifier $\kappa$ and a basetype $\beta$, neither of which may be omitted.
Further, $\beta$ is a type without a type qualifier, but which has type qualifiers on all its type parameters.
Formally, $\beta ::= C|C \angles{\tau}$ where $C$ represents a Java basetype.

@@

\todo{A type $\tau$ is represented as $\kappa \ \beta$
where $\kappa$ is a type qualifier (that is, \<NonDet>, \<OrderNonDet>, or
\<Det>) and $\beta$ is a Java basetype.}

@@

\todo{Note that the \rulename{ordernondet} versions also handle \<Det>,
  because $\<Det> \sqsubseteq \<OrderNonDet>$.}

@@

\todo{\todo{I think this will be OK, by redefining $\beta$.}
  The collection rules do not permit nested collections; similarly, the
  (currently commented out) array rules did not permit any two-dimensional
  array to be well-formed.  There is a bigger issue here.
  Consider the ``nondet collection'' rule.  $\beta_c$ is a class (we need
  to use a different symbol for it, probably $c$), but $\beta_e$ should be
  a full type, which might contain type qualifiers within it.  (Actually,
  $\|NonDet|\ \beta_e$ is a type, not $\beta_e$ itself.)  So, there should
  be an antecedent stating that $\|NonDet| \beta_e$ is well-formed, and
  for the ``ordernondet collection'' rule stating that $\kappa_e \beta_e$
  is well-formed.  Likewise, antecedents are needed for the map rules.
  Otherwise, these rules permit construction of $\<Map>\angles{\tau'}$
  where $\tau'$ is itself not well-defined.}

@@

The element type of a \<NonDet> collection must be \<NonDet>.
Otherwise, the type qualifier on a collection's element must be a subtype or equal to
the collection's type qualifier (\rulename{collection}\ rules of
\cref{type-validity}).
\Cref{fig:determinism-collections} restates its restrictions, and here are examples.

>>>>

Figure~\cref{typecheck-rules-standard} shows a sample of the standard typing rules applied to 
any typed language. In the generic case, the type $\tau$ represents a valid basetype in a programming language.
\ourTypeSystem extends a basetype with a type qualifier. Throughout this paper, $\tau$ represents a fully qualified type.
In other words, type $\tau = \kappa \ \beta$ consists of both a qualifier $\kappa$ and a basetype $\beta$, neither of which may be omitted.
Further, $\beta$ is a type without a type qualifier, but which has type qualifiers on all its type parameters.
Formally, $\beta ::= C|C \angles{\tau}$ where $C$ represents a valid Java basetype.
\ourTypeSystem checks all the standard typing rules including the ones shown in Figure~\cref{typecheck-rules-standard} with
the only difference of having an additional type qualifier on each of the basetypes. The type qualifiers can only 
constrain the set of valid types. As a result, \ourTypeSystem doesn't allow any behavior that the original type system doesn't.

@@

%\todo{A type $\tau$ is represented as $\kappa \ \beta$
%where $\kappa$ is a type qualifier (that is, \<NonDet>, \<OrderNonDet>, or
%\<Det>) and $\beta$ is a Java basetype.}

@@

%\todo{Note that the \rulename{ordernondet} versions also handle \<Det>,
%  because $\<Det> \sqsubseteq \<OrderNonDet>$.}

@@

%\todo{\todo{I think this will be OK, by redefining $\beta$.}
%  The collection rules do not permit nested collections; similarly, the
%  (currently commented out) array rules did not permit any two-dimensional
%  array to be well-formed.  There is a bigger issue here.
%  Consider the ``nondet collection'' rule.  $\beta_c$ is a class (we need
%  to use a different symbol for it, probably $c$), but $\beta_e$ should be
%  a full type, which might contain type qualifiers within it.  (Actually,
%  $\|NonDet|\ \beta_e$ is a type, not $\beta_e$ itself.)  So, there should
%  be an antecedent stating that $\|NonDet| \beta_e$ is well-formed, and
%  for the ``ordernondet collection'' rule stating that $\kappa_e \beta_e$
%  is well-formed.  Likewise, antecedents are needed for the map rules.
%  Otherwise, these rules permit construction of $\<Map>\angles{\tau'}$
%  where $\tau'$ is itself not well-defined.}

@@

The element type of a \<NonDet> collection must be \<NonDet>.
Otherwise, the type qualifier on a collection's element must be a subtype or equal to
the collection's type qualifier (\rulename{collection}\ rules of
\cref{type-validity}). Note that the \rulename{ordernondet} versions also handle \<Det>,
because $\<Det> \sqsubseteq \<OrderNonDet>$.
\Cref{fig:determinism-collections} restates its restrictions, and here are examples.
